---
title: "Creating ECS input files for presentation/analysis of data from a surface-rupturing earthquake"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example-1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(ECSprep)
suppressPackageStartupMessages(library(dplyr)) 

# Note that dplyr is required for ECSprep.
# It required here for piping inside user functions.
```

## Objective

Presentation and analysis of post-earthquake fault rupture and displacement
data, where the user has mapped traces of surface ruptures and locations where
fault displacement was measured.

This example shows how to use `ECSprep` to create the a ruptures input file and
a fault displacement measurements input file for the Lavrentiadis & Abrahamson 
ECS algorithm from standard Esri shapefiles..

## Example data

For this example, we will use data from the 1915 Pleasant Valley earthquake in
Nevada. The original data are from [Wallace et al. (1984)](https://doi.org/10.3133/pp1274AB)
and are included in the FDHI Project Database Flatfiles ([Sarmiento et al., 2021](https://doi.org/10.34948/N3JS31))
as Earthquake #34. 

The shapefiles are provided in `inst/extdata` in this package.

## Preparing the faults

The Lavrentiadis & Abrahamson Event Coordinate System (ECS) algorithm develops
a nominal fault system trace from mapped faults represented as ordered vertices
for each trace.

Just like the example in [Creating ECS input files for Fault Displacement Hazard Analysis](https://github.com/asarmy/ECSprep/blob/master/vignettes/example-1.Rmd), 
we will use the `create_ecs_input_ruptures.R` function of this package.

It takes one required argument, `in_filepath`, and one optional argument for the 
`output_folder` where the CSV will be saved. 

- The `in_filepath` must provide an Esri shapefile. Other formats are not 
supported.

- If an `output_folder` is not specified, the CSV will be saved in the same 
directory as the input shapefile.

The CSV will retain all the attributes in the input shapefile. It will also add
the following attributes:

- `RUP_ID` is a unique identifier for each fault trace line.

- `NODE_ID` is the order of the vertices for each fault trace line.

- `Longitude` is the X coordinate in WGS84 (EPSG:4326) for the NODE_ID.

- `Latitude` is the Y coordinate in WGS84 (EPSG:4326) for the NODE_ID.

If the `RUP_ID` column exists in the input shapefile, it will retained as-is.

If there is not an attribute for `Rank`, then it will be added automatically and 
populated with "None." It is recommended that users assign  *Principal* and 
*Distributed* ranks because the ECS algorithm works better with rankings. 

In our example here, the shapefile contained a `Rank` column with valid ranks.

```{r}
input_shp <- system.file("extdata", "pleasant_valley_final_ruptures.shp", package="ECSprep")
create_ecs_input_ruptures(input_shp)
```

We can import the CSV and confirm that the new columns were added and the original
ones were retained:

```{r}
file_out <- system.file("extdata", "ECSprep-outputs", "rupture_vertices-pleasant_valley_final_ruptures.csv", package="ECSprep")
result <-  read.csv(file.path(file_out))
names(result)
```

This CSV is can be used as the ruptures input for the Lavrentiadis & Abrahamson 
ECS algorithm tool.


## Preparing the fault displacement measurement sites

Most fault displacement hazard analysis models for principal faulting use the
along-strike location of the site as a predictor variable. Similarly, most 
models for distributed faulting use the strike-normal distance of the site as
a predictor variable.

The Lavrentiadis & Abrahamson algorithm will can use the data to influence the 
ECS in addition to the impact the fault traces have on the ECS.

We will use the `create_ecs_input_measurements.R` function of this package.

It takes one required argument, `in_filepath`, and one optional argument for the 
`output_folder` where the CSV will be saved. 

Note that a column called `displ` that contains the fault displacement
measurements is required. As long as all values are in the same units, the unit
does not matter because only the relative amplitude is important.

- The `in_filepath` must provide an Esri shapefile. Other formats are not 
supported.

- If a `displ_column` is not specified, then the user will be notified and the 
script will halt execution.

- If an `output_folder` is not specified, the CSV will be saved in the same 
directory as the input shapefile.

The CSV will retain all the attributes in the input shapefile. It will also add
the following attributes:

- `PT_ID` is a unique identifier for each site .

- `Longitude` is the X coordinate in WGS84 (EPSG:4326) for the SITE_ID.

- `Latitude` is the Y coordinate in WGS84 (EPSG:4326) for the SITE_ID.

If the `PT_ID` column exists in the input shapefile, it will retained as-is.

If there is not an attribute for `Rank`, then it will be added automatically and 
populated with "None." It is recommended that users assign  *Principal* and 
*Distributed* ranks because the ECS algorithm works better with rankings. 

In our example here, the shapefile contained a `Rank` column with valid ranks.

```{r}
input_shp <- system.file("extdata", "pleasant_valley_final_measurements.shp", package="ECSprep")
create_ecs_input_measurements(input_shp)
```

We can import the CSV and inspect the contents:

```{r}
file_out <- system.file("extdata", "ECSprep-outputs", "displacement_sites-pleasant_valley_final_measurements.csv", package="ECSprep")
result <-  read.csv(file.path(file_out))

options(width = 120)
head(result, 5)
```


This CSV is can be used as an optional fault displacement measurement input for 
the Lavrentiadis & Abrahamson ECS algorithm tool.

*Note this file is easy to create manually* as the only required contents are
columns for `PT_ID`, `displ`, Latitude`, and `Longitude`.
