---
title: "Creating ECS input files for Fault Displacement Hazard Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example-1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(ECSprep)
suppressPackageStartupMessages(library(dplyr)) 

# Note that dplyr is required for ECSprep.
# It required here for piping inside user functions.
```

## Objective

Forward-application of fault displacement hazard analysis models, where the
user has mapped traces of active faults and site locations for hazard analysis.

This example shows how to use ECSprep to create the ruptures input files for
the Lavrentiadis & Abrahamson ECS algorithm from standard Esri shapefiles.

## Example data

For this example, we will use the trace of the Newport-Inglewood/Rose Canyon
Fault System from the US Geological Survey Quarternary Fault and Fold Database. 
It is provided as an Esri shapefile in `inst/extdata` in this package.

We will also consider a randomly chosen site along the NI/RC. It is saved as
an Esri shapefile in `inst/extdata` in this package.

## Preparing the faults

The Lavrentiadis & Abrahamson Event Coordinate System (ECS) algorithm develops
a nominal fault system trace from mapped faults represented as ordered vertices
for each trace.

We will use the `create_ecs_input_ruptures.R` function of this package.

It takes one required argument, `in_filepath`, and one optional argument for the 
`output_folder` where the CSV will be saved. 

- The `in_filepath` must provide an Esri shapefile. Other formats are not 
supported.

- If an `output_folder` is not specified, the CSV will be saved in the same 
directory as the input shapefile.

The CSV will retain all the attributes in the input shapefile. It will also add
the following attributes:

- `RUP_ID` is a unique identifier for each fault trace line.

- `NODE_ID` is the order of the vertices for each fault trace line.

- `Longitude` is the X coordinate in WGS84 (EPSG:4326) for the NODE_ID.

- `Latitude` is the Y coordinate in WGS84 (EPSG:4326) for the NODE_ID.

If the `RUP_ID` column exists in the input shapefile, it will retained as-is.

If there is not an attribute for `Rank`, then it will be added automatically and 
populated with "None." It is recommended that users assign  *Principal* and 
*Distributed* ranks because the ECS algorithm works better with rankings. 

In our example here, the shapefile did not contain ranks, so we get a warning 
that this column was added. If the `Rank` column contains invalid values, then 
the user will be notified and the script will halt execution.

```{r}
input_shp <- system.file("extdata", "usgs_qffd_nirc.shp", package="ECSprep")
create_ecs_input_ruptures(input_shp)
```

We can import the CSV and confirm that the new columns were added and the original
ones were retained:

```{r}
file_out <- system.file("extdata","ECSprep-outputs", "rupture_vertices-usgs_qffd_nirc.csv", package="ECSprep")
result <-  read.csv(file.path(file_out))
names(result)
```

This CSV is can be used as the ruptures input for the Lavrentiadis & Abrahamson 
ECS algorithm tool.


## Preparing the hazard sites

Most fault displacement hazard analysis models for principal faulting use the
along-strike location of the site as a predictor variable. Similarly, most 
models for distributed faulting use the strike-normal distance of the site as
a predictor variable.

The Lavrentiadis & Abrahamson algorithm can provide the ECS ordinates of a 
hazard site if the ECS is established from the mapped faults (and optionally
fault displacement measurement locations).

We will use the `create_ecs_input_haz_sites.R` function of this package.

It takes one required argument, `in_filepath`, and one optional argument for the 
`output_folder` where the CSV will be saved. 

- The `in_filepath` must provide an Esri shapefile. Other formats are not 
supported.

- If an `output_folder` is not specified, the CSV will be saved in the same 
directory as the input shapefile.

The CSV will retain all the attributes in the input shapefile. It will also add
the following attributes:

- `SITE_ID` is a unique identifier for each site .

- `Longitude` is the X coordinate in WGS84 (EPSG:4326) for the SITE_ID.

- `Latitude` is the Y coordinate in WGS84 (EPSG:4326) for the SITE_ID.


```{r}
input_shp <- system.file("extdata", "example_site.shp", package="ECSprep")
create_ecs_input_haz_sites(input_shp)
```

We can import the CSV and inspect the contents:

```{r}
file_out <- system.file("extdata", "ECSprep-outputs", "hazard_sites-example_site.csv", package="ECSprep")
result <-  read.csv(file.path(file_out))
result
```


This CSV is can be used as an optional hazard site input for the Lavrentiadis & 
Abrahamson ECS algorithm tool.

*Note this file is easy to create manually* as the only required contents are
columns for `SITE_ID`, `Latitude`, and `Longitude`.
