---
title: "Creating ECS input files for Fault Displacement Hazard Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example-1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(ECSprep)
suppressPackageStartupMessages(library(dplyr)) 

# Note that dplyr is required for ECSprep.
# It required here for piping inside user functions.
```

The Lavrentiadis & Abrahamson Event Coordinate System (ECS) algorithm develops
a nominal fault system trace from mapped faults represented as ordered vertices
for each trace.

This example shows how to use `ECSprep` to create the ruptures input file for
the Lavrentiadis & Abrahamson ECS algorithm from a standard Esri shapefile.

For this example, we will use the trace of the Newport-Inglewood Fault from
the US Geological Survey Quarternary Fault and Fold Database. It is provided in
`inst/extdata`.

We will use the main user function of this package, which is `create_ecs_input.R`.
It takes two required ordered arguments: `in_filepath` and `geometry_type` and
optionally the `output_folder` for the CSV. 

- The `in_filepath` must provide an Esri shapefile. Other formats are not supported.

- The `geometry_type` will be "line" for our example here, which uses fault traces.
(The other use case is "point" when fault displacement measurement sites are used."

- If an `output_folder` is not specified, the CSV will be saved in the same 
directory as the input shapefile.

The CSV will retain all the attributes in the input shapefile. It will also add
the following attributes:

- `RUP_ID` is a unique identifier for each fault trace line.

- `NODE_ID` is the order of the vertices for each fault trace line.

- `Longitude` is the X coordinate in WGS84 (EPSG:4326) for the NODE_ID.

- `Latitude` is the Y coordinate in WGS84 (EPSG:4326) for the NODE_ID.

If there is not an attribute for `Rank`, then it will be added automatically and 
populated with "None." It is recommended that users assign  *Principal* and 
*Distributed* ranks because the ECS algorithm works better with rankings. In our 
example here, the shapefile did not contain ranks, so we get a warning that this 
column was added. If the `Rank` column contains invalid values, then the user
will be notified.

```{r}
input_shp <- system.file("extdata", "usgs_qffd_nirc.shp", package="ECSprep")
create_ecs_input(input_shp, "line")
```

We can import the CSV and confirm that the new columns were added and the original
ones were retained:

```{r}
file_out <- system.file("extdata","ECSprep-outputs", "rupture_vertices-usgs_qffd_nirc.csv", package="ECSprep")
result <-  read.csv(file.path(file_out))
names(result)
```

This CSV is can be used as the ruptures input for the Lavrentiadis & Abrahamson 
ECS algorithm tool.
